---
title: "Final report"
author: "Marlon E. Cobos"
date: "May 1, 2018"
output: pdf_document
csl: Science.csl
bibliography: My Library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r checkpoint, echo=FALSE, cache=TRUE, warning=FALSE, message=FALSE}
if(!require(checkpoint)){ #if require install this package
  install.packages("checkpoint")
  library(checkpoint)
}
dir.create(file.path(tempdir(), ".checkpoint"), recursive = TRUE, showWarnings = FALSE)
checkpoint("2018-04-30", checkpointLocation = tempdir())
```

```{r packages, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, results='hide'}
#Packages
packages <- c("ku.enm", "maptools", "rgdal", "rgeos", "rasterVis", "raster",
              "classInt", "RColorBrewer")
toInstall <- packages[!packages %in% installed.packages()]
if(length(toInstall)>0L) install.packages(toInstall)
lapply(packages, require, character.only = TRUE)
```

## Introduction

Amphibians are among the most threatened species of vertebrates and their conservation status has continue declining during the last decades @blaustein_direct_2010. One of the main concerns for future conservation of these species is Climate Change @li_review_2013. This threat may cause direct or indirect impacts, affecting these species phenology, physiology, and distributional patterns @blaustein_complexity_2011. Insular amphibians are specially vulnerable to climate change-related threats because of the particular topographic and climatic characteristics of archipelagos @bellard_vulnerability_2014. 

*Peltophryne empusa* is one the eight endemic Cuban toads, its current conservation status is Vulnerable owing to its reduced and fragmented area of occupancy @iucn_peltophryne_2004-6. Previous studies have revealed that Cuban amphibians my be under risk because of future changes in climate @cobos_recent_2016. The main potential effect of this phenomenon on these toads that has been predicted is a reduction in their distributional area. Previous evaluations of this species conservation status have not considered potential threats from climate change @iucn_peltophryne_2004-6; hence a new evaluation that considers this potential changes to the future is necessary.

The aim of this project is to re-assess the conservation status of *P. empusa* considering potential effects of climate change on its area of occupancy. To do that ecological niche modeling techniques were used to reconstruct this species current and future potential distribution. Based on these results, a reassessment of its conservation status was performed following the IUCN criteria @iucn_iucn_2014 and considering how losses of suitability would affect the area of occupancy of this species.    

## Methods

#### Study site

The study area is the entire Cuban archipelago (Figure 1). This insular country is part of the Great Antilles and its geological origin and history is one of the more complex of the world @formel_constitucion_1989.

```{r study_area, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=6, fig.align="center", fig.cap="Study area, model clibration area, and species occurrences.", cache=TRUE}
# data
#cuba <- readGDAL("")
#m <- readGDAL("")
occ <- read.csv("pemp_joint.csv")

# plot


```


#### Species and environmental data

A total of 246 species occurrences were obtained from online databases (GBIF and VertNet), the scientific literature @alonso_bosch_origen_2011, and data from field expeditions. A spatial thinning process was performed to reducing auto-correlation and a total of 64 records remained. These records were split by randomly selecting 25% of them for testing and the remaining for training models.

Bioclimatic variables from the WorldClim database @hijmans_very_2005 were used as environmental predictors for the present scenario (resolution ~1km). For representing future scenarios bioclimatic variables of the representative concentration pathway (RCP) 4.5, for the NCAR-CCSM4 general circulation model (GCM; @stocker_climate_2013) were used. Future data was obtained from the CGIAR Research Program on Climate Change, Agriculture and Food Security database (CCAFS; available at http://www.ccafs-climate.org/data_spatial_downscaling/). A process of jackknife was performed in Maxent to select three candidate sets of variables that contribute the most to the model and that were not more than 80% correlated with each other.

#### Ecological niche model's calibration

A total of 150 candidate models were evaluated as part of the model calibration process. These models were created in Maxent with distinct parameter settings resulted from combining 3 sets of environmental predictors, 10 regularization multipliers, and 5 feature classes.

```{r candidate_models, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, results='hide'}
#Variables with information to be used as arguments
occ_joint <- "pemp_joint.csv"
occ_tra <- "pemp_tra.csv"
M_var_dir <- "M"
batch_cal <- "emp_enm_calibration_models"
cal_dir <- "Calibration_Models"
reg_mul <- c(seq(0.5,1,0.1),seq(2,5,1))
fea_cs <- "basic"
running <- TRUE
```

```{r candidate_models1, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, results='hide',eval=FALSE}
ku.enm.cal(occ.joint = occ_joint, occ.tra = occ_tra, M.var.dir = M_var_dir, batch = batch_cal,
           out.dir = cal_dir, reg.mult = reg_mul, f.clas = fea_cs, run = running)
```

A process of evaluation was performed for selecting candidate models and their associated parameters. Optimal models were selected based on three distinct criteria: statistical significance (based on partial ROC; @peterson_rethinking_2008), prediction ability (omission rate allowing an error *E* = 5%), and model complexity (AICc). 

```{r evaluation_selection, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, results='hide'}
#Variables with information to be used as arguments
occ_test <- "pemp_test.csv"
out_eval <- "emp_enm_evaluation_results"
ov <- 5
rp <- 50
ni <- 100
kt <- TRUE
selc <- "OR_AICc"
##Some of the variables used here as arguments were already created for the previous function
```

```{r evaluation_selection1, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, results='hide',eval=FALSE}
ku.enm.ceval(path = cal_dir, occ.joint = occ_joint, occ.tra = occ_tra, occ.test = occ_test, batch = batch_cal,
            out.eval = out_eval, omi.val = ov, rand.perc = rp, no.inter = ni, kept = kt, selection = selc)
```

#### Final model creation

Final models were created using the parameters of the best candidate models, inside the calibration area, performing five replicates by Bootstrap, and using the complete sets of occurrences. These models were transferred to all the Cuban archipelago and the future scenario (i.e., RCP 4.5) in the selected GCM. Model extrapolations were not allowed to avoid obtaining predictions in areas with non-analogous climatic conditions.

```{r preliminar_models, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, results='hide'}
#Variables with information to be used as arguments
bat_fin <- "Final_models"
mod_dir <- "Final_Models"
G_var_dir <- "G"
rep <- 5
rtyp <- "Bootstrap"
ofor <- "logistic"
pr <- TRUE
extyp <- "no_ext"
jack <- FALSE
##Most of the variables used here as arguments were already created for the previous function
```

```{r preliminar_models1, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, results='hide',eval=FALSE}
ku.enm.mod(occ.joint = occ_joint, M.var.dir = M_var_dir, out.eval = out_eval, batch = bat_fin, rep.n = rep, rep.type = rtyp, 
           jackknife = jack, out.dir = mod_dir, out.format = ofor, project = pr, G.var.dir = G_var_dir, ext.type = extyp)
```

#### Species conservation status re-evaluation

Continues outputs of final models (current and future) were summarized by calculating the median of their median results. These medians where binarized to obtain suitable and unsuitable areas for the species using the minimum value of suitability in the species occurrences as a threshold. Changes between current and future suitable areas were categorized as stable, lost, and gained areas. The change of suitability was assessed for all known populations of this toad and for areas of occupancy of this species considered in previous assessments @iucn_peltophryne_2004-6. The species conservation status was re-evaluated according to the B2ab(iii) criteria from the International Union for Conservation of Nature @iucn_iucn_2014.  

```{r model_analysis, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE}
# medians of final models
mod1 <- raster("Final_Models/M_0.7_F_lqpth_Set2_NE/Peltophryne_empusa_Cuba_median.asc")
mod2 <- raster("Final_Models/M_0.8_F_lqpth_Set2_NE/Peltophryne_empusa_Cuba_median.asc")

modf1 <- raster("Final_Models/M_0.7_F_lqpth_Set2_NE/Peltophryne_empusa_ccsm4_4.5_median.asc")
modf2 <- raster("Final_Models/M_0.8_F_lqpth_Set2_NE/Peltophryne_empusa_ccsm4_4.5_median.asc")

med <- calc(stack(mod1, mod2), median)
medf <- calc(stack(modf1, modf2), median)

# binarization of models
suit_val <- extract(med, occ[, 2:3])
thres <- sort(suit_val)[1] # threshold based on a 3% error value
threshold <- cbind(c(0, thres), c(thres, 1), c(0, 1))
thresholdf <- cbind(c(0, thres), c(thres, 1), c(0, 2))

bin <- reclassify(med, threshold)
binf <- reclassify(medf, thresholdf)

bina <- na.omit(getValues(bin))
areabin <- length(subset(bina, bina == 1))

# evaluating impact of climate change
## changes in distributional areas
changes <- calc(stack(bin, binf), sum)

area <- na.omit(getValues(changes))
lost_area <- length(subset(area, area == 1))
gain_area <- length(subset(area, area == 2))
stable_parea <- length(subset(area, area == 3))

## populations and future suitability
suit_chang <- extract(changes, occ[, 2:3])
occ$Suit_change <- suit_chang

## re-evaluation of conservation status
#pe_iucn <- readGDAL("")
#changes_in <- mask(changes, pe_iucn)

#areas <- na.omit(getValues(changes_in))
#lost_pe_iucn <- length(subset(areas, areas == 1))
```

## Results

```{r results1, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, results='hide'}
ress <- read.csv(paste(out_eval, "evaluation_stats.csv", sep = "/"))
resb <- read.csv(paste(out_eval, "best_models_OR_AICc.csv", sep = "/"))
resc <- read.csv(paste(out_eval, "evaluation_results.csv", sep = "/"))
```

A total of `r ress[1,1]` of `r length(resc[,1])` candidate models were significant. Since none of the candidate models resulted in omission rates lower than 5%, `r length(resb[,1])` best parameters were chosen by their AICc values (delta AICc < 2, calculated among models with the lowest omission rate).

```{r results2, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=4, fig.align="center", fig.cap="Distribution of all, non-statistically significant, and selected models according to their omission rates and AICc values.", cache=TRUE}
knitr::include_graphics(paste(out_eval, "evaluation_figure", sep = "/"))
```


Species distribution probabilities were higher in lowlands. Highlands in the archipelago did not present suitable conditions for the species. Considering these results and the general trend in climate for this region to the future (higher temperature and lower precipitation), environmental suitability may increase towards the mountains. However, whether the environments in lowlands will be more or less suitable in the future is uncertain. This is why future model projections are one the main results to get in future analyses. Comparisons of current and future Boolean models derived from logistic outputs of Maxent are other results that will allow estimating the area

```{r results3, echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=6, fig.align="center", fig.cap="Preliminar geographic representation of the ecological niche model and ocurrences for the studied species.", cache=TRUE}
# ploting
#colfunc <- colorRampPalette(c("red","yellow","royalblue","navyblue"))
plot(changes, col = c("grey95", "red", "blue", "grey70"))
points(occ$Longitud, occ$Latitud, col = "black", cex = 0.7, pch = 16)

```


## Discussion

This study results revealed that some populations of this toad could be affected owing to the lost of climatic suitability in places where they are currently distributed. The area of occupancy of the species, consequently, would be smaller, specially if migration to new suitable areas is discarded. However, an important factor to be considered is that suitable areas generally over-predict occupied areas, since they represent adequate environmental conditions for the species and not its presence. When   

## References