---
title: "Tests"
author: "Marlon E. Cobos"
date: "March 31, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages and functions

```{r packages_functions, echo=T, cache=TRUE}
#Packages
packages <- c("rmarkdown", "sqldf", "knitr", "ENMeval", "yaml")
toInstall <- packages[!packages %in% installed.packages()]
if(length(toInstall)>0L) install.packages(toInstall)
lapply(packages, require, character.only = TRUE)

#Functions
source("https://raw.githubusercontent.com/manubio13/ME_Cobos_FProject/master/cal_function.R")
source("https://raw.githubusercontent.com/manubio13/ME_Cobos_FProject/master/eval_function.R")
source("https://raw.githubusercontent.com/manubio13/ME_Cobos_FProject/master/mod_function.R")
```


## Tests for ku.enm.cal

Test if it can create a batch file for running maxent models

```{r ku.enm.cal_test_1, echo=T, cache=TRUE}
#Variables with information to be used as arguments
occ_all <- "pemp_all.csv"
occ_tra <- "pemp_tra.csv"
M_var_dir <- "M"
batch_cal <- "emp_enm_calibration_models1"
cal_dir <- "Calibration_Models1"
reg_mul <- c(seq(0.5,1,0.1),seq(2,5,1))
fea_cs <- "basic"
running <- FALSE

#Executing the function
ku.enm.cal(occ.all = occ_all, occ.tra = occ_tra, M.var.dir = M_var_dir, batch = batch_cal,
           out.dir = cal_dir, reg.mult = reg_mul, f.clas = fea_cs, run = running)

if (exists("emp_enm_calibration_models1.bat"))
{
  print("passed")
} else
{
  print("failed")
}
```

Test if it runs the models automatically if run = TRUE. VISUAL TEST

```{r ku.enm.cal_test_2, echo=T, cache=TRUE}
#Variables with information to be used as arguments
running <- TRUE

#Executing the function
ku.enm.cal(occ.all = occ_all, occ.tra = occ_tra, M.var.dir = M_var_dir, batch = batch_cal,
           out.dir = cal_dir, reg.mult = reg_mul, f.clas = fea_cs, run = running)
```

Test if batch length is appropriate acording to feature classes selection

```{r ku.enm.cal_test_3, echo=T, cache=TRUE}
batch_cal <- "emp_enm_calibration_models2"
running <- FALSE

#Executing the function
ku.enm.cal(occ.all = occ_all, occ.tra = occ_tra, M.var.dir = M_var_dir, batch = batch_cal,
           out.dir = cal_dir, reg.mult = reg_mul, f.clas = fea_cs, run = running)

#Number of models expected to result with the defined varaibles
n <- 150
bf <- readLines("emp_enm_calibration_models2.bat") 
nr <- length(bf)

if (nr == n)
{
  print("passed")
} else
{
  print("failed")
}
```


# Tests for ku.enm.eval

Test if it creates the folder for the evaluation results

```{r ku.enm.eval_test_1, echo=T, cache=TRUE}
#Variables with information to be used as arguments
occ_test <- "pemp_test.csv"
out_eval <- "emp_enm_evaluation_results"
ov <- 5
rp <- 50
ni <- 100
kt <- TRUE
selc <- "OR_AICc"
##Most of the variables used here as arguments were already created for the previous function

#Executing the function
ku.enm.eval(path = cal_dir, occ.all = occ_all, occ.tra = occ_tra, occ.test = occ_test, batch = batch_cal,
            out.eval = out_eval, omi.val = ov, rand.perc = rp, no.inter = ni, kept = kt, selection = selc)

dr <- dir()

if (out_eval %in% dr)
{
  print("passed")
} else
{
  print("failed")
}
```

Test if all evaluation files were created

```{r ku.enm.eval_test_2, echo=T, cache=TRUE}
#Files that should exist
files <- c ("best_models_OR_AICc.csv", "evaluation_figure.png", "evaluation_results.csv", "evaluation_results.html", "evaluation_stats.csv")

ev_files <- list.files(path = out_eval)

if (files %in% ev_files)
{
  print("passed")
} else
{
  print("failed")
}
```


# Tests for ku.enm.mod

Test if folder for final models was created 

```{r ku.enm.mod_test_1, echo=T, cache=TRUE}
#Variables with information to be used as arguments
mod_dir <- "Final_Models"
G_var_dir <- "G_variables"
rep <- 5
rtyp <- "Bootstrap"
ofor <- "logistic"
pr <- TRUE
extyp <- "all"
##Most of the variables used here as arguments were already created for the previous function

#Executing the function
ku.enm.mod(occ.all = occ_all, M.var.dir = M_var_dir, out.eval = out_eval, rep.n = rep, rep.type = rtyp, 
          out.dir = mod_dir, out.format = ofor, project = pr, G.var.dir = G_var_dir, ext.type = extyp)

dr1 <- dir()

if (G_var_dir %in% dr1)
{
  print("passed")
} else
{
  print("failed")
}
```

Test if all subdirectories were created for final models in the G_var_dir folder

```{r ku.enm.cal_test_2, echo=T, cache=TRUE}
#Subdirectories that should exist
dirn <- read.csv(paste(out_eval, "best_models_OR_AICc.csv", sep = "/"))[,1]
ext_op <- c("E", "EC", "NE")
sdirs <- paste(rep(dirn, 3), ext_op, sep = "_")

fm_dir <- list.filesdir(path = G_var_dir)

if (sdirs == fm_dir)
{
  print("passed")
} else
{
  print("failed")
}
```

