---
title: "Tests"
author: "Marlon E. Cobos"
date: "March 31, 2018"
output: pdf_document
---

<!--***DAN: Search for ***DAN in all files for my comments. All files may have them.-->

<!--***DAN: The scope is huge, this is certainly a real research project, and if all you plan is accomplished, it will be more than enough for a final project! I hope the computing tools and approaches you have leared are helping! 

<!--***DAN: Your git use is good in the sense that you have done many commits with sensible messages and used tags. I hope it has been a useful tool!-->

<!--***DAN: I cannot evaluate the code because I am not expert in this area, and there is a lot of it, obviously written over a span of time. This is one reason why unit tests can be so helpful! I am glad to see you have implemented them according to an appropriate framework!-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages and functions

```{r packages_functions, echo=TRUE, cache=TRUE, include=TRUE}
#Packages
packages <- c("rmarkdown", "sqldf", "knitr", "ENMeval", "yaml")
toInstall <- packages[!packages %in% installed.packages()]
if(length(toInstall)>0L) install.packages(toInstall)
lapply(packages, require, character.only = TRUE)

#Functions
source("https://raw.githubusercontent.com/manubio13/ME_Cobos_FProject/master/cal_function.R")
source("https://raw.githubusercontent.com/manubio13/ME_Cobos_FProject/master/eval_function.R")
source("https://raw.githubusercontent.com/manubio13/ME_Cobos_FProject/master/mod_function.R")
```


## Tests for cal_function

Test if it can create a batch file for running maxent models

```{r cal_test_1, echo=TRUE, cache=TRUE, include=TRUE}
#Variables with information to be used as arguments
occ_all <- "pemp_all.csv"
occ_tra <- "pemp_tra.csv"
M_var_dir <- "M"
batch_cal <- "emp_enm_calibration_models1"
cal_dir <- "Calibration_Models1"
reg_mul <- c(seq(0.5,1,0.1),seq(2,5,1))
fea_cs <- "basic"
running <- TRUE

#Executing the function
ku.enm.cal(occ.all = occ_all, occ.tra = occ_tra, M.var.dir = M_var_dir, batch = batch_cal,
           out.dir = cal_dir, reg.mult = reg_mul, f.clas = fea_cs, run = running)

if (file.exists(paste(batch_cal, "bat", sep = "."))) {
  cat("\npassed")
} else {
  cat("\nfailed")
}
```

Test if batch length is appropriate acording to feature classes selection

```{r cal_test_2, echo=TRUE, cache=TRUE, include=TRUE}
#Number of models expected to result with the defined varaibles
if(fea_cs == "all" | fea_cs == "basic" | fea_cs == "no.t.h" | 
   fea_cs == "no.t.h" | fea_cs == "no.t.h"){
  if(fea_cs == "all"){
    n <- prod(length(reg_mul), length(dir("M")), 29, 2) #number of potential 
    #combinations times two (two models for each combination are being created)
  }
  if(fea_cs == "basic"){
    n <- prod(length(reg_mul), length(dir("M")), 5, 2) 
  }
  if(fea_cs == "no.t.h"){
    n <- prod(length(reg_mul), length(dir("M")), 7, 2) 
  }
  if(fea_cs == "no.t"){
    n <- prod(length(reg_mul), length(dir("M")), 15, 2) 
  }
  if(fea_cs == "no.h"){
    n <- prod(length(reg_mul), length(dir("M")), 15, 2) 
  }
}else{
  n <- length(fea_cs)
}

bf <- readLines(paste(batch_cal, "bat", sep = ".")) 
nr <- length(bf)

if (nr == n) {
  cat("\npassed")
} else {
  cat("\nfailed")
}
```


# Tests for eval_function

Test if it creates the folder for the evaluation results

```{r eval_test_1, echo=TRUE, cache=TRUE, include=TRUE}
#Variables with information to be used as arguments
occ_test <- "pemp_test.csv"
out_eval <- "emp_enm_evaluation_results1"
ov <- 5
rp <- 50
ni <- 100
kt <- TRUE
selc <- "OR_AICc"
##Most of the variables used here as arguments were already created 
```

```{r eval_test_1a, echo=TRUE, cache=TRUE, include=TRUE, eval=FALSE}
#Executing the function
ku.enm.eval(path = cal_dir, occ.all = occ_all, occ.tra = occ_tra, occ.test = occ_test, 
            batch = batch_cal, out.eval = out_eval, omi.val = ov, rand.perc = rp, 
            no.inter = ni, kept = kt, selection = selc)
```

```{r eval_test_1b, echo=TRUE, cache=TRUE, include=TRUE}
dr <- dir()

if (out_eval %in% dr){
  cat("\npassed")
} else{
  cat("\nfailed")
}
```

Test if all evaluation files were created

```{r eval_test_2, echo=TRUE, cache=TRUE, include=TRUE}
#Files that should exist
files <- c ("best_models_OR_AICc.csv", "evaluation_figure.png", "evaluation_results.csv", 
            "evaluation_results.html", "evaluation_stats.csv")

ev_files <- list.files(path = out_eval)

if (sum(files == ev_files) == 5){
  cat("\npassed")
} else{
  cat("\nfailed")
}
```


# Tests for mod_function

Test if folder for final models was created 

```{r mod_test_1, echo=TRUE, cache=TRUE, include=TRUE}
#Variables with information to be used as arguments
mod_dir <- "Final_Models1"
G_var_dir <- "G"
rep <- 5
rtyp <- "Bootstrap"
ofor <- "logistic"
pr <- TRUE
extyp <- "all"
##Most of the variables used here as arguments were already created 

#Executing the function
ku.enm.mod(occ.all = occ_all, M.var.dir = M_var_dir, out.eval = out_eval, 
           rep.n = rep, rep.type = rtyp, out.dir = mod_dir, out.format = ofor, 
           project = pr, G.var.dir = G_var_dir, ext.type = extyp)

dr1 <- dir()

if (G_var_dir %in% dr1) {
  cat("\npassed")
} else{
  cat("\nfailed")
}
```

Test if all subdirectories were created for final models in the Final models folder

```{r mod_test_2, echo=TRUE, cache=TRUE, include=TRUE}
#Subdirectories that should exist
dirn <- read.csv(paste(out_eval, "best_models_OR_AICc.csv", sep = "/"))[,1]
ext_op <- c("E", "EC", "NE")
sdirs <- paste(rep(dirn, 3), ext_op, sep = "_")

fm_dir <- list.files(path = mod_dir)

if (sum(sdirs %in% fm_dir) == length(sdirs)){
  cat("\npassed")
} else{
  cat("\nfailed")
}
```

